---
title: "Survival Analysis and Cox Regression"
output: HTML
  word_document: default
  html_document: default
date: '2022-07-25'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Survival Analysis
A survival analysis would be carried out to show how long on average a baby is exclusively breastfed

### Data Description
duplicate_data : duplicated dataset from the primary questionnaire data before childbirth
breastfeed_fllw : follow up data on breastfeeding after childbirth
surv_data : merged dataset of the primary and follow up data
surv: optimised dataset for survival and cox analysis

```{r}
library(tidyverse)
library(visdat)
library(dplyr)
library(lubridate)
# load follow up data
breastfeed_fllw <- read_csv("~/Documents/PMIM202J/Assignment data/Born in Wales Data June 2022 2/Born in Wales Follow-up Data June 2022.csv")

duplicate_data <- read.csv("~/Documents/PMIM202J/Assignment data/Born in Wales Data June 2022 2/Born in Wales Data June 2022.csv")

head(breastfeed_fllw) # preview first lines of data
str(breastfeed_fllw) # format variables

```

# Merge primary data with follow up data
```{r}
# rename columns in duplicated dataset
colnames(duplicate_data) <- c('ID', 'start_time', 'expected_del_date', 'language', 'ethnicity', 'nationality', 'relationship', 'sexuality', 'covid_symp', 'symp_start', 'symp_type', 'covid_treat', 'test', 'low_mood', 'interest', 'depression', 'sleep_stat', 'tiredness', 'eating', 'down_feeling', 'trouble', 'movement', 'suicidal', 'smoke', 'alcohol_intake', 'anxiety', 'worry_ctrl', 'worry_too_much', 'trouble_relaxing', 'restlessness', 'annoyed_irritable', 'scared', 'pa_freq', 'pa', 'cycling', 'walking', 'home_childcare', 'gardening', 'walk_pace', 'weight', 'height', 'preg_stress', 'rel_issues', 'legal_financial_issues', 'victim_of_crime', 'serious_illness', 'death', 'covid_related', 'emotion_financial_support', 'stress_range', 'healthcare_range', 'covid_preg_change', 'maternity_care', 'midwife_support', 'baby_feed', 'other_children', 'no_children', 'child_age_gap', 'occupation', 'education', 'work', 'work_basis', 'no_occupants', 'income', 'talk_to_outsiders', 'contact_with_outsiders', 'answers_count', 'wimd_2019_rank', 'wimd_2019_decile', 'wimd_2019_quintile', 'wimd_2019_quantile', 'covid_vaccine_during_preg', 'submit')

head(duplicate_data) # preview first lines of data
# notice there are empty values, change these empty values to NA so they can be easy to analyse
require(dplyr)
duplicate_data <- duplicate_data %>% mutate_all(na_if, "")

# rename columns in follow up dataset
colnames(breastfeed_fllw) <- c('ID', 'Start_time', 'delivery_date', 'started_breastfeeding', 'stopped_breastfeeding')

# merge follow up data and primary data by ID
surv_data <- merge(duplicate_data, breastfeed_fllw, by=c("ID"))
```


```{r}
# create censor for observations without a stopped breastfeeding observation
# This covers mothers that started breastfeeding but were lost during the study
surv_data$censored <- ifelse(is.na(surv_data$stopped_breastfeeding), 0, 1)

# create a new column that creates the censored end of study date 
for (i in 1:nrow(surv_data)) {if(is.na(surv_data$stopped_breastfeeding[i])) 
    surv_data$end_date[i] <- as.character(surv_data$delivery_date[i] %m+% years(1))
                        else
                            surv_data$end_date[i] <- as.character(surv_data$stopped_breastfeeding[i])
}
surv_data$end_date <- as.Date(surv_data$end_date, format = "%Y-%m-%d")

# create a new column to display the time to event (length of breastfeeding in days) for each respondents
surv_data$LOS <- as.Date(surv_data$end_date) - as.Date(surv_data$started_breastfeeding)

# recode and group features for survival analysis and cox regression

# emotional and financial support
surv_data <- surv_data %>% mutate(emotion_financial_support = recode(emotion_financial_support,
                                                                       "Yes" = "yes",
                                                                       "No" = "no",
                                                                       "No " = "no"))

surv_data <- surv_data %>% mutate(emotion_financial_support=ifelse(emotion_financial_support == "yes", 0, 1)) #convert to binary

surv_data$emotion_financial_support <- as.factor(surv_data$emotion_financial_support) # convert to factor


# work based demand level
surv_data <- surv_data %>% mutate(pa_freq = recode(pa_freq, "I spend most of my time at work sitting (such as in an office)" = "gentle", 
                                                   "My work involves definite physical effort including handling of heavy objects and use of tools (e.g. cleaner, hospital nurse, gardener, postal delivery workers etc.)" = "intense", 
                                                   "I spend most of my time at work standing or walking. However, my work does not require much intense physical effort (e.g. shop assistant, hairdresser, security guard, childminder, etc.)" = "moderate", 
                                                   "I am not in employment (e.g. retired, retired for health reasons, unemployed, full-time carer etc.)" = "gentle", 
                                                   "My work involves vigorous physical activity including handling of very heavy objects (e.g. construction worker, refuse collector, etc.)" = "intense", 
                                                   "I am not in employment (e.g. stay at home parent, unemployed, full-time carer etc.)" = "gentle"))

surv_data <- surv_data %>% mutate(pa_freq = ifelse(pa_freq == "gentle", 0, ifelse(pa_freq == "moderate", 1, 2))) # convert to binary
surv_data$pa_freq <- as.factor(surv_data$pa_freq) # convert to factor

# select features needed for survival analysis, we do not need the 'ID' and 'stopped_breastfeeding' since we already have the end date feature
surv <- select(surv_data, c("Start_time", "started_breastfeeding", "end_date", "censored", "LOS", "pa_freq", "emotion_financial_support"))
view(surv)

# remove missing observations
surv <- na.omit(surv)

# One last thing, rename the columns before survival analysis
colnames(surv) <- c('start_time', 'started_breastfeeding', 'end_date', 'censored', 'LOS', 'workload', 'domestic_support')
```

### Survival Analysis
```{r}
# load the libraries for survival analysis
# install.packages("survival")
# install.packages("survminer")
library(survival)
library(survminer)

# create a table to display stopped breastfeeding and censored by domestic support table
table2 <- table(surv$domestic_support, surv$censored) 
row.names(table2) <- c("domestic_support=yes", "domestic_support=no")
colnames(table2) <- c("stopped breastfeeding", "censored")
table2
```

### Overall breastfeeding duration
```{r}
attach(surv)
gen <- Surv(LOS, censored) # create survival object for breastfeeding time to event
head(gen, 20)

gen_fit <- survfit(gen ~ 1, data = surv)
summary(gen_fit, times = c(1, 30, 60*(1:20))) # restrict time to event to every 30 days

# plot support survival fit
gen_plot <- ggsurvplot(gen_fit, data = surv, 
           # change labels and titles
           ylab="Survival Probability", xlab="Time (days)",
           legend.labs=c("All women"),
           legend.title="",
           surv.scale="percent",
           title="Survival of Breastfeeding",
           subtitle = "By Kaplan Meier",
           # change fonts and colours
           font.title = c(20, "bold", "black"),
           font.subtitle = c(10, "black"),
           font.x=c(10, "bold"),
           font.y=c(10, "bold"),
           font.xtickslab=c(10, "plain"),
           font.ytickslab=c(10, "plain"),
           palette = c("purple"),
           # adjust scaling
           surv.plot.height = 0.65,
           break.x.by = 30,
           censor = TRUE,
           censor.shape="|",
           censor.size=2,
           tables.col="strata",
           # confidence interval
           conf.int.alpha=c(0.1),
           conf.int = TRUE,
           # add risk table
           risk.table = TRUE,
           risk.table.height = 0.35,
           risk.table.fontsize = 3.0,
           risk.table.col = "strata")
```



### Breastfeeding duration by domestic support
```{r}
# create a survival object for domestic support
# domestic support
attach(surv)
fit_support <- survfit(Surv(LOS, censored)~domestic_support) # create survival object for breastfeeding time to event by domestic support

print(fit_support)
summary(fit_support)

summary(fit_support, times = c(1, 30, 60*(1:20))) # restrict time to event to every 30 days

# plot support survival fit
support_plot <- ggsurvplot(fit_support, data = surv, 
           # change labels and titles
           ylab="Survival Probability", xlab="Time (days)",
           legend.labs=c("received support", "did not receive support"),
           legend.title="",
           surv.scale="percent",
           title="Survival of Breastfeeding",
           subtitle = "By Domestic Support (Kaplan Meier)",
           # change fonts and colours
           font.title = c(20, "bold", "black"),
           font.subtitle = c(10, "black"),
           font.x=c(10, "bold"),
           font.y=c(10, "bold"),
           font.xtickslab=c(10, "plain"),
           font.ytickslab=c(10, "plain"),
           palette = c("purple", "brown"),
           # adjust scaling
           surv.plot.height = 0.65,
           break.x.by = 30,
           censor = TRUE,
           censor.shape="|",
           censor.size=2,
           tables.col="strata",
           # confidence interval
           conf.int.alpha=c(0.1),
           conf.int = TRUE,
           # add risk table
           risk.table = TRUE,
           risk.table.height = 0.35,
           risk.table.fontsize = 3.0,
           risk.table.col = "strata",
           #add p-value
           pval = TRUE,
           pval.size = 4,
           pval.coord = c(1,0,1))

```
There is a steady decline in stopping breastfeeding in women that received domestic support than women who did not. A sharp descent can be seen when stopping breastfeeding for women who did not receive domestic support.


### Breastfeeding duration by workload survival analysis

```{r}
# workload

# stopped breastfeeding by workload table
table3 <- table(surv$workload, surv$censored) 
unique(surv$workload)
row.names(table3) <- c("workload=gentle", "workload=moderate", "workload=intense")
colnames(table3) <- c("stopped breastfeeding", "censored")
table3

# Survival Analysis
# create survival object for workload
attach(surv)
fit_workload <- survfit(Surv(LOS, censored)~workload)
print(fit_workload)
summary(fit_workload)

summary(fit_workload, times = c(1, 30, 60*(1:20))) # restrict time to event to every 30 days


workload_plot <- ggsurvplot(fit_workload, data = surv, 
           # change labels and titles
           ylab="Survival Probability", xlab="Time (days)",
           legend.labs=c("gentle", "moderate", "intense"),
           legend.title="",
           surv.scale="percent",
           title="Survival of Breastfeeding",
           subtitle = "By Workload (Kaplan Meier Plot)",
           # change styles, fonts and colours
           font.title = c(20, "bold", "black"),
           font.subtitle = c(10, "black"),
           font.x=c(10, "bold"),
           font.y=c(10, "bold"),
           font.xtickslab=c(10, "plain"),
           font.ytickslab=c(10, "plain"),
           # adjust scaling
           surv.plot.height = 0.65,
           break.x.by = 30,
           censor = TRUE,
           censor.shape="|",
           censor.size=2,
           tables.col="strata",
           # confidence interval
           conf.int.alpha=c(0.1),
           conf.int = TRUE,
           # add risk table
           risk.table = TRUE,
           risk.table.height = 0.35,
           risk.table.fontsize = 3.0,
           risk.table.col = "strata",
           #add p-value
           pval = TRUE,
           pval.size = 4,
           pval.coord = c(1,0,1))
```
- There is a higher survival rate in women that undergo intense workload to stopping breastfeeding than women who undergo moderate and gentle workload.

### Cox Regression
```{r}
# create cox model for workload and domestic support
cox_model1 <- coxph(Surv(LOS, censored)~workload + domestic_support, data=surv)
summary(cox_model1)

# plot cox model
cox_fit <- survfit(cox_model1)
cox_plot <- ggsurvplot(cox_fit, data = surv, 
           main="cox model", 
           xlab="Time (days)",
           surv.scale="percent",
           title="Survival of Breastfeeding by Domestic Support and Workload (Cox Plot)")

```


```{r}
# create cox model for workload
cox_workload <- coxph(Surv(LOS, censored)~workload, data=surv)
summary(cox_workload)

# plot cox workload model
cox_fit2 <- survfit(cox_workload)
ggsurvplot(cox_fit, data = surv, 
           main="cox model", 
           xlab="Time (days)",
           surv.scale="percent",
           title="Survival of Breastfeeding by Workload (Cox Plot)")

```

```{r}

# create cox model for support
cox_support <- coxph(Surv(LOS, censored)~domestic_support, data=surv)
summary(cox_support)

# plot cox domestic support model
cox_fit3 <- survfit(cox_support)
ggsurvplot(cox_fit, data = surv, 
           main="cox model", 
           xlab="Time (days)",
           surv.scale="percent",
           title="Survival of Breastfeeding by Domestic Support (Cox Plot)")
```


```{r}
# Hazard Ratio
ggforest(cox_model1, data = surv)

```

### Check for a smooth fitted curve to know if there are variations overtime
```{r}
# difference between survival curves vy each feature
# workload
survdiff(Surv(LOS, censored)~workload, data = surv)

#support
survdiff(Surv(LOS, censored)~domestic_support, data = surv)
```

### The Proportional Hazards Assumption of a Cox Regression can be tested using cox.zph
```{r}
# test for a hazard ratio
cox_tos <- coxph(Surv(LOS, censored)~domestic_support + workload, data=surv)
zph_test1 <- cox.zph(cox_tos)
print(zph_test1)
```
According to the results, the test is not statistically significant for any of the covariates, and the global test is not statistically significant either. From this result, we would say our proportional hazards assumption is sound.

### Look for smooth fitted curves
```{r}
# A graphical way to view proportional hazards for individual variables
ggcoxzph(zph_test1)
```
- The fitted curve is smooth and this indicates they do not vary overtime.

```{r}
sessionInfo()
```

